Improving reproducibility of scientific software using Nix/NixOS: A case study on preCICE adapters and solvers

### 2024-05-10 paper/paper.tex
fix(paper): Expand all short formulations of ' and replace "which"

preCICE is an open-source project, which means its source code is freely available and can be modified and distributed by the community.
This fosters collaboration and encourages the development of new features and improvements by researchers and engineers worldwide.

@@

To see how well Nix can be used to build and package scientific software reliably and reproducibly, we conduct a case study in the context of the latest preCICE distribution~\cite{preciceDistribution}, which is currently preCICE \texttt{v2211.0}.
The preCICE distribution consists out of the preCICE core library, as well as some tools, bindings, adapters, tutorials, the website and documentation, and the preCICE VM\footnote{\url{https://precice.org/installation-distribution.html}}.
Table \ref{table:label-distribution} shows an overview over all packages provided by the preCICE distribution and their respected versions.

@@

This package is a collection of tools which help in development of new adapters.
It stands for Artificial Solver Testing Environment, so it allows for debugging adapter mapping setups and other similar tasks.

ASTE\footnote{\url{https://github.com/precice/aste}} needs VTK9\footnote{\url{https://vtk.org/}}, a visualization toolkit, which is built without python support in the nixpkgs repository to reduce compilation time.
However, this feature is needed for ASTE.
As already mentioned, Nix allows easy overriding of inputs, so in this case, we can look at the package definition of VTK.
The parameter \texttt{enablePython} can simply be set to \texttt{true}, then we also need to supply the version of python we want to compile against, so in our case \texttt{python = python3;}.

@@

A preCICE distribution comes bundled with several adapters, which are currently not in the upstream nixpkgs repository.
Many of the adapters use solvers that aren't packaged in the upstream repo as well, so in order to get the adapters to work, these also have to be packaged and sometimes patched.

@@

The preCICE adapter for CSM code CalculiX~\cite{Uekermann2017_Adapters}, requires the original CalculiX source code and provides a new Makefile in the adapter repository.
By default, this Makefile expects the source code to reside in \texttt{\$HOME}, but it is possible to override this location with a make variable.
It then builds the original code and the adapter code together, which results in a combined binary.

The Makefile also has hard-coded dependency location for \texttt{SPOOLES}, \texttt{ARPACK} and \texttt{YAML} so these need to be replaced with the equivalent \texttt{pkg-config} calls, because Nix doesn't have these dependencies at the hard-coded locations, and generally only finds dependencies with \texttt{pkg-config}.
Additionally, there is no \texttt{install} target provided by the Makefile so the resulting binary needs to be installed manually by copying it to \texttt{\$out}.\\

@@

With this approach, we disable HDF5, which is already packaged in the upstream nixpkgs repository, even at the required version 5.1.10.
HDF5 is packaged with multiple outputs.
One package, \texttt{out}, contains the libraries and binaries and another, \texttt{dev} contains the header files, this is, because most of the time, users installing HDF5 only need the libraries and binaries, and it helps to reduce installation size of this rather large package.
Additionally, code\_aster requires medfile and scotch, which are both available upstream and can be used as a dependency.

Another dependency, which is also provided by code\_aster is metis, which partially builds, so we do not replace this dependency, but we need to fix the second part of the build, which includes the metis programs such as gpmetis, ndmetis, mpmetis and more.
The issue is that in the \texttt{CMakeLists.txt} for the metis programs \texttt{link\_directories} is hard-coded to \texttt{/home/karypis/local/lib}.
After removing this line, it also installs, so to resolve this issue in the buildPhase, we unpack the metis tarball, remove this line with \texttt{sed} and then repack the directory.

@@

Mumps builds using a Makefile and can be configured using Makefile.inc, by default it provides a couple of these configurations for systems such as Ubuntu, so we use one of these provided configuration files to write our own custom config file.
It defines the location for dependencies such as scotch, metis, parmetis, liblapack, blas, scalapack and blacs.
Almost all the dependencies, are available upstream, but we need to recompile scotch with additional build flags, \texttt{scotch ptscotch esmumps ptesmumps} using Nix overrideAttrs feature.
For packaging blacs, which also uses a Makefile, we configure the build using a custom \texttt{Bmake.inc} file, which is used to set compiler flags, the path to the mpi and bash installation.

The last missing dependency for code\_aster, tfel, uses CMake as a build system which makes packaging it trivial with Nix and we only have to set a couple of CMake flags.

After successfully packaging all dependencies, the build phase of code\_aster completes, but the installation part has another issue, that also affects new versions of Ubuntu and other distributions which use a python version greater than 3.9.
There is a forum post without any resolution from the developers or the community, so we need to manually patch the bug.

The issue is that the custom build system does not correctly calculate the python site-packages directory because it unconditionally slices the first 3 chars from \texttt{sys.version}, which works if your version is 3.9.x but not with 3.10.x because then the code calculates \texttt{lib64/python3.1/site-packages} rather than the expected \texttt{lib64/python3.10/site-packages}.

@@

Last but not least, we can copy the code\_aster adapter at the required location which concludes this adapter installation.\\

@@

In order to package this adapter, we first need to package deal.II~\cite{dealII95}, which uses CMake as build system and works without any adjustments.
The same applies for the adapter, which also uses CMake and needs preCICE as a dependency.
When supplied with the inputs, the adapter builds without any problems.
As an example for the Nix language, you can see the abbreviated Nix code for the dealii.II adapter in Fig. \ref{lst:dealii-adapter-nix}.\\

@@

The DUNE~\cite{bastian2020dune} package uses a combination of CMake files and custom build scripts which makes the build process quite tedious.
This might be an issue because of the ongoing migration to CMake from autotools and might be resolved in the future.

@@

For the build and install, process we need to manually set \texttt{\$DUNE\_CONTROL\_PATH} and \texttt{\$DUNE\_PY\_DIR} environment variables, both variables also need to be set at runtime.
Additionally, we need to patch the python install process because the current CMake file that handles this install tries to access the internet with \texttt{pip install} which isn't allowed in Nix' sandboxed builds.
Therefore, we need to patch this \texttt{pip} command with the options \texttt{--no-build-isolation --no-cache-dir --no-index --no-deps} and we need to remove the \texttt{--upgrade} option.\\

@@

We are explicitly referring to the legacy version of FEniCS~\cite{fenics} here, not the newer and still experimental FEniCSx.
This adapter~\cite{Rodenberg2021} uses the open source computing platform FEniCS, which is already packaged within nixpkgs, to provide preCICE support.

Although it is available, not all features the adapter needs are enabled, including PETSc support.
To fix this, we need to package PETSc4py, which provides python bindings for PETSc and is not present in the upstream nixpkgs repository.
The build process is possible without any issues, because it uses the internal \texttt{buildPythonApplication} build tool, but we need to add the additional build optional \texttt{build\_src --force} because the cython code needs to be rebuilt.
We also aren't able to enable tests for this package because tests depend on OpenMPI and the network, which isn't available within the Nix build sandbox.

@@

Building the adapter with Nix is now possible with \texttt{buildPythonPackage} and doesn't require any special workarounds.
A majority of the \texttt{pytest} ran successfully, but two tests failed inside the Nix sandbox, which means all tests needed to be disabled.\\

@@

OpenFOAM uses its own custom build system called \texttt{wmake} which is sometimes called with a wrapper script called \texttt{Allwmake}.
The build system sets 36 environment variables that are set by wmake, one step at a time by checking several parameters, e.g. the cpu architecture or the location of the source code, and concatenating some other environment variables together.

Allwmake and wmake rely on these environment variables to be set, as otherwise the tools will exit with a warning.
During runtime, before running any OpenFOAM commands, a small wrapper script has to be started that sets all relevant environment variables, or a user has to manually source a file inside the installation directory of OpenFOAM.
OpenFOAM uses modules for extensibility and flexibility which need the source code of OpenFOAM and also require wmake as a build tool.

For Nix, these properties are rather unfortunate.
To compile OpenFOAM with Nix, we first need to patch the shebangs\footnote{\url{https://foldoc.org/shebang}} of wmake to make it run during the build.
Secondly, we render a shell script we call \texttt{set-openfoam-vars} that exports all necessary environment variables to the current shell session.
At the time of writing, all parameters such as the processor architecture are hard-coded for simplicity reasons but could be parametrized based on the Nix inputs to allow for optimized builds.
During the buildPhase, the rendered script will be sourced to set all variables, e.g. so the \texttt{OPENFOAM\_SRC\_PATH} points to \texttt{/build/openfoam} which is the default location for Nix builds inside the sandbox.
After sourcing the script, the single command \texttt{./Allwmake -j -q} is sufficient to start the build.

@@

OpenFOAM is available on Spack and EasyBuild which both patch the OpenFOAM source files as those tools also have to set environment variables to make a build compatible with wmake.\\

@@

This package is a preCICE adapter~\cite{Uekermann2017_Adapters} for the CFD code SU2 and extends the original SU2 source code by patching the files.
For this, the adapter provides a custom script that needs to be executed before the build phase.
Nix offers the patchPhase which allows us to run the script in this phase and patch the original source, after that the \texttt{stdenv} automatically recognizes that the repository uses autotools for building the software and runs the configurePhase.

@@

Generally speaking, there are major gaps regarding the ability to successfully package the software between each different software solution that can be traced back to the build system.
Custom build systems require an disproportional amount of additional work to make them runnable on any system.
OpenFOAM and Allwmake are prime examples of these issues because now package managers have to understand a whole new build system for exactly one package in order to successfully package it and make it easily available to users.
It is also worth mentioning that writing your own custom build system, requires additional time to maintain this build system, which could have been spent working on the software itself.
Packages that use common, industry proven package managers, such as CMake and autotools make it easier for developers of the software to build their software as well as package managers to package it.
These build tools, generally also provide interfaces for dependency management which can be used to tell users and package maintainers which dependencies are needed, and it is usually better than only documenting required dependencies, because documentation needs to be kept in sync with the underlying code.
They also generally already provide proven interfaces for configuring packages with additional features, and these also might be self documenting.

The authors of the xSDK\footnote{\url{https://xsdk.info/}} came to a very similar conclusion.
Their goal is to ``identify, adapt, and adopt best practices in software engineering''~\cite{xsdk-website} and ``to provide the foundation of this extensible scientific software ecosystem''~\cite{xsdk-website}.
To achieve this they provide a list of package policies~\cite{xSDK2023}, which for example specifies as mandatory policy that packages ``must support portable installation through Spack''~\cite{xSDK2023} and that all packages ``should have a build system that is appropriate for the language''~\cite{xSDK2023}, which includes CMake and autoconf as examples.
If a package supports all these policies, they can be added to the growing list of packages in the xSDK, which currently among other things include PETSc, deal.ii and preCICE.
These packages are already part of nixpkgs, or it is easy to provide a package for them because they support portable installation, meaning installation into custom prefixes such as nix store paths, and common build systems such as CMake and autoconf.

@@

This chapter compares Nix with two widely used solution in HPC environments, Spack and EasyBuild.
There was a similar comparison done at FOSDEM 2018~\cite{comparision-fosdem} which also included conda and Guix, but this was by the release manager of EasyBuild and we want to provide an update with this comparison.
The comparison first takes a high level look on some statistics and how they compare and after that we will discuss, usability, reproducibly and performance in more detail.

@@

The first table~\ref{table:comparison-overview} shows some basic stats about the amount of packages the solutions provide out of the box and how they are implemented.
This is important because the higher the number the more likely it is that your package is already packaged and doesn't require additional work.
It is also important to look how packages are implemented because, if a package needs to be packaged by for example researches, domain specific languages have the downside that, researches now have to get familiar with this DSL, while general purpose languages such as python do not have this downside because they are already used in their field.
Although this is a downside, DSL usually are design with a purpose in mind and can get you faster to the desired outcome.

@@

Spack's way of installing has some problems, with using \texttt{git clone} on the latest default branch, in their case \texttt{develop}, this has the implication, that you aren't using the latest stable release of the software but rather a random commit, that might have broken something recently.
It also has the problem that you can't easily update the package repository independent from the package manager, this requires additional work from the users if they require this.

It is also worth mentioning that all 3 solutions do not require root for installation but Nix has the option for a system wide installation of the Nix daemon, which then can be used to install packages locally into the users PATH, which no longer requires root.

@@

The basic usage of all three tools is pretty similar and they support searching and installing package in a simple manner, all this is show in table~\ref{table:comparison-usage}.
The difference in installing software comes down to the unique way that Nix installs software, it downloads/builds the paths into its local store, which can either be located in the user's home directory or in the path \texttt{/nix/store} depending on the installation.
\texttt{nix-shell} later only puts these downloaded paths into the \texttt{\$PATH} environment variable.
Spack and EasyBuild on the other hand use the widely used \texttt{module} system that usually is already found on HPC clusters, it allows loading software into the current environment.

Installing old software also differs for all three solutions, Nix on the one hand keeps some old, widely used software packages around and maintains those, examples are \texttt{boost} which is available in different versions or e.g. \texttt{hdf5} which is available in the latest version and version \texttt{1.10}.
The other way of installing older software on nix is using something that is called \texttt{overrideAttrs}, it can be used to change anything of a package, such as the used sources, meaning it can be used to replace it with an older/newer version or it can be used to add an patch, add a new dependency or install additional files into the \texttt{\$out} directory.
An example would be the following \texttt{nix-shell -p 'precice.overrideAttrs(\_: \{ version = "2.4.0"; src = \ldots \} )'}, it replaces the precice version with 2.4.0 and adds the old source, which will then trigger a new build.
Spack and EasyBuild on the other hand, keep the old software versions in the specifications or keep the whole specification, so you can reinstall an old version by specifying the available version, for example, \texttt{spack install gcc@6.4.0}
If the required version is not available, you have to edit the current package definition and add the required version.

To achieve reproducibility, it is necessary that we can install software from a previous revision of the packages.
Nix allows us to simply specify the channel revision we want to use to build our software, it then evaluates the requested packages and builds it.
This can easily be done without affecting the current system by specifying it in the shell command \texttt{nix-shell -p pan -I nixpkgs=pathToRevision/urlToTarball}, it now only uses this channel for building the requested package.
Spack, can achieve something similar by checking out your Spack version at a previous revision and then installing the package requested package, but this approach has a downside that it affects the whole Spack installation, and now also work with an older Spack tool.
On the other hand, EasyBuild, could be installed using pip at an earlier version, but it also offers something different called toolchains, which we will look at shortly.

Next, we look at how you can build a package with a specific compiler version using specific compiler flags.
The flexibility of Nix overrides allows different levels of rebuilds of packages with a specific compiler, you can either just rebuild the requested package with specific compiler, this for example rebuilds curl with clang8 \texttt{nix-shell -p '(curl.override \{ stdenv = clang8Stdenv;\})'} or you can rebuild the package and all required dependencies with a specific compiler by overriding the \texttt{stdenv} on nixpkgs level, \texttt{nix-build -E 'with (import <nixpkgs> \{ config.replaceStdenv = \{ pkgs \}: pkgs.clang8Stdenv; \}); curl'}.
This is possible because every c/c++/\ldots package, is build using the \texttt{stdenv} which defines the compiler and default compile flags.
While Nix provides a default stdenv, which is used to compile all packages, it is possible to override and replace the package with your another stdenv provided by nixpkgs, such as the clang8Stdenv, you can write your own stdenv, or you can extend the current stdenv with flags.
By default, there are some useful helpers provided, such as \texttt{makeStaticLibraries} which allows compiling any library as a static library, or \texttt{enableDebugging} which sets the compiler flags \texttt{-ggdb -Og} and can then be enabled on the package plus all dependencies.
Spack provides an easy way to specify the compiler using \texttt{spack install precice \%clang} and only requires a registered compiler, it can be a system compiler, or the compiler can be installed using Spack.
On this note, Spack tooling and interface to override compilers, flags and more is way more user friendly and readable than using and invoking the Nix DSL within a one line command call.
EasyBuild has a whole different system, where packages have defined toolchains.\footnote{\url{https://docs.easybuild.io/common-toolchains/}}
A toolchain is a set of packages defined at a specific version and users can create their own toolchains.
For example, EasyBuild provides a foss toolchain which contains packages such as, binutils, gcc, OpenMPI, openblas and more and packages are provided against a specific toolchain version such as foss-2022b.
Packages are then provided with toolchains in mind and are never deleted, so installing a package with a different compiler can be achieved by selecting a different eb script with a different toolchain version.
On the other hand this results in a less elegant user flow, because if a user wants to install a package with a toolchain that isn't provided they have to fork a package definition and update the toolchain.\\

@@

Spack does something similar to Nix where it adds hashes to each installation path, but also stores additional files inside the package installation which include the complete spec of the package than can later be used to reproducibly build the same package.~\cite{Gamblin_2015}.
It also does something that Nix, currently doesn't support, which is it storing the architecture and cpu features that are used in this build, that way it can provide binary packages that are optimized for specific CPUs.
None the less it has one problem that Nix doesn't have: it allows packages to link back into system dependencies, for example, while building hpl-2.3 using a gcc 12.2.0 compiled with Spack we see the usage of the following system dependencies \texttt{libm.so}, \texttt{libc.so}, \texttt{libdl.so}, \texttt{libnl-3.so}, \texttt{libnl-route-3.so}, \texttt{librt.so}, \texttt{libutil.so}, \texttt{libpthread.so} as well as the dynamic loader.
Most of these dependencies are stable, and usually do not change on a system update, but it also means that it simply can't guarantee the best possible reproducibility and it gets worse if the system compiler was used rather than a compiler build from scratch using Spack.
It is also worth mentioning that the Spack developers are aware of this problem and are currently working on building these system libraries themselves~\footnote{\url{https://github.com/spack/spack/issues/39560}}~\footnote{\url{https://github.com/spack/spack/pull/42082}}.

>>>>

preCICE is an open-source project, that means its source code is freely available and can be modified and distributed by the community.
This fosters collaboration and encourages the development of new features and improvements by researchers and engineers worldwide.

@@

To see how well Nix can be used to build and package scientific software reliably and reproducibly, we conduct a case study in the context of the latest preCICE distribution~\cite{preciceDistribution}, currently preCICE \texttt{v2211.0}.
The preCICE distribution consists out of the preCICE core library, as well as some tools, bindings, adapters, tutorials, the website and documentation, and the preCICE VM\footnote{\url{https://precice.org/installation-distribution.html}}.
Table \ref{table:label-distribution} shows an overview over all packages provided by the preCICE distribution and their respected versions.

@@

This package is a collection of tools that help in development of new adapters.
It stands for Artificial Solver Testing Environment, so it allows for debugging adapter mapping setups and other similar tasks.

ASTE\footnote{\url{https://github.com/precice/aste}} needs VTK9\footnote{\url{https://vtk.org/}}, a visualization toolkit, that is built without python support in the nixpkgs repository to reduce compilation time.
However, this feature is needed for ASTE.
As already mentioned, Nix allows easy overriding of inputs, so in this case, we can look at the package definition of VTK.
The parameter \texttt{enablePython} can simply be set to \texttt{true}, then we also need to supply the version of python we want to compile against, so in our case \texttt{python = python3;}.

@@

A preCICE distribution comes bundled with several adapters, that are currently not in the upstream nixpkgs repository.
Many of the adapters use solvers that are not packaged in the upstream repo as well, so in order to get the adapters to work, these also have to be packaged and sometimes patched.

@@

The preCICE adapter for CSM code CalculiX~\cite{Uekermann2017_Adapters}, requires the original CalculiX source code and provides a new Makefile in the adapter repository.
By default, this Makefile expects the source code to reside in \texttt{\$HOME}, but it is possible to override this location with a make variable.
It then builds the original code and the adapter code together, resulting in a combined binary.

The Makefile also has hard-coded dependency location for \texttt{SPOOLES}, \texttt{ARPACK} and \texttt{YAML} so these need to be replaced with the equivalent \texttt{pkg-config} calls, because Nix does not have these dependencies at the hard-coded locations, and generally only finds dependencies with \texttt{pkg-config}.
Additionally, there is no \texttt{install} target provided by the Makefile so the resulting binary needs to be installed manually by copying it to \texttt{\$out}.\\

@@

With this approach, we disable HDF5, as it is already packaged in the upstream nixpkgs repository, even at the required version 5.1.10.
HDF5 is packaged with multiple outputs.
One package, \texttt{out}, contains the libraries and binaries and another, \texttt{dev} contains the header files, this is, because most of the time, users installing HDF5 only need the libraries and binaries, and it helps to reduce installation size of this rather large package.
Additionally, code\_aster requires medfile and scotch, that are both available upstream and can be used as a dependency.

Another dependency, that is also provided by code\_aster is metis, that partially builds, so we do not replace this dependency, but we need to fix the second part of the build, including the metis programs such as gpmetis, ndmetis, mpmetis and more.
The issue is that in the \texttt{CMakeLists.txt} for the metis programs \texttt{link\_directories} is hard-coded to \texttt{/home/karypis/local/lib}.
After removing this line, it also installs, so to resolve this issue in the buildPhase, we unpack the metis tarball, remove this line with \texttt{sed} and then repack the directory.

@@

Mumps builds using a Makefile and can be configured using Makefile.inc, by default it provides a couple of these configurations for systems such as Ubuntu, so we use one of these provided configuration files to write our own custom config file.
It defines the location for dependencies such as scotch, metis, parmetis, liblapack, blas, scalapack and blacs.
Almost all the dependencies, are available upstream, but we need to recompile scotch with additional build flags, \texttt{scotch ptscotch esmumps ptesmumps} using Nix overrideAttrs feature.
For packaging blacs, that also uses a Makefile, we configure the build using a custom \texttt{Bmake.inc} file, that is used to set compiler flags, the path to the mpi and bash installation.

The last missing dependency for code\_aster, tfel, uses CMake as a build system making it trivial to package it with Nix and we only have to set a couple of CMake flags.

After successfully packaging all dependencies, the build phase of code\_aster completes, but the installation part has another issue, that also affects new versions of Ubuntu and other distributions that use a python version greater than 3.9.
There is a forum post without any resolution from the developers or the community, so we need to manually patch the bug.

The issue is that the custom build system does not correctly calculate the python site-packages directory because it unconditionally slices the first 3 chars from \texttt{sys.version}, that works if your version is 3.9.x but not with 3.10.x because then the code calculates \texttt{lib64/python3.1/site-packages} rather than the expected \texttt{lib64/python3.10/site-packages}.

@@

Last but not least, we can copy the code\_aster adapter at the required location.\\

@@

In order to package this adapter, we first need to package deal.II~\cite{dealII95}, that uses CMake as build system and works without any adjustments.
The same applies for the adapter, that also uses CMake and needs preCICE as a dependency.
When supplied with the inputs, the adapter builds without any problems.
As an example for the Nix language, you can see the abbreviated Nix code for the dealii.II adapter in Fig. \ref{lst:dealii-adapter-nix}.\\

@@

The DUNE~\cite{bastian2020dune} package uses a combination of CMake files and custom build scripts making the build process quite tedious.
This might be an issue because of the ongoing migration to CMake from autotools and might be resolved in the future.

@@

For the build and install, process we need to manually set \texttt{\$DUNE\_CONTROL\_PATH} and \texttt{\$DUNE\_PY\_DIR} environment variables, both variables also need to be set at runtime.
Additionally, we need to patch the python install process because the current CMake file that handles this install tries to access the internet with \texttt{pip install}.
This is not allowed in Nix' sandboxed builds.
Therefore, we need to patch this \texttt{pip} command with the options \texttt{--no-build-isolation --no-cache-dir --no-index --no-deps} and we need to remove the \texttt{--upgrade} option.\\

@@

We are explicitly referring to the legacy version of FEniCS~\cite{fenics} here, not the newer and still experimental FEniCSx.
This adapter~\cite{Rodenberg2021} uses the open source computing platform FEniCS, that is already packaged within nixpkgs, to provide preCICE support.

Although it is available, not all features the adapter needs are enabled, including PETSc support.
To fix this, we need to package PETSc4py, that provides python bindings for PETSc and is not present in the upstream nixpkgs repository.
The build process is possible without any issues, because it uses the internal \texttt{buildPythonApplication} build tool, but we need to add the additional build optional \texttt{build\_src --force} because the cython code needs to be rebuilt.
We also are not able to enable tests for this package because tests depend on OpenMPI and the network, that is not available within the Nix build sandbox.

@@

Building the adapter with Nix is now possible with \texttt{buildPythonPackage} and does not require any special workarounds.
A majority of the \texttt{pytest} ran successfully, but two tests failed inside the Nix sandbox, that means all tests needed to be disabled.\\

@@

OpenFOAM uses its own custom build system called \texttt{wmake} that is sometimes called with a wrapper script called \texttt{Allwmake}.
The build system sets 36 environment variables that are set by wmake, one step at a time by checking several parameters, e.g. the cpu architecture or the location of the source code, and concatenating some other environment variables together.

Allwmake and wmake rely on these environment variables to be set, as otherwise the tools will exit with a warning.
During runtime, before running any OpenFOAM commands, a small wrapper script has to be started that sets all relevant environment variables, or a user has to manually source a file inside the installation directory of OpenFOAM.
OpenFOAM uses modules for extensibility and flexibility that need the source code of OpenFOAM and also require wmake as a build tool.

For Nix, these properties are rather unfortunate.
To compile OpenFOAM with Nix, we first need to patch the shebangs\footnote{\url{https://foldoc.org/shebang}} of wmake to make it run during the build.
Secondly, we render a shell script we call \texttt{set-openfoam-vars} that exports all necessary environment variables to the current shell session.
At the time of writing, all parameters such as the processor architecture are hard-coded for simplicity reasons but could be parametrized based on the Nix inputs to allow for optimized builds.
During the buildPhase, the rendered script will be sourced to set all variables, e.g. so the \texttt{OPENFOAM\_SRC\_PATH} points to \texttt{/build/openfoam} being the default location for Nix builds inside the sandbox.
After sourcing the script, the single command \texttt{./Allwmake -j -q} is sufficient to start the build.

@@

OpenFOAM is available on Spack and EasyBuild that both patch the OpenFOAM source files as those tools also have to set environment variables to make a build compatible with wmake.\\

@@

This package is a preCICE adapter~\cite{Uekermann2017_Adapters} for the CFD code SU2 and extends the original SU2 source code by patching the files.
For this, the adapter provides a custom script that needs to be executed before the build phase.
Nix offers the patchPhase that allows us to run the script in this phase and patch the original source, after that the \texttt{stdenv} automatically recognizes that the repository uses autotools for building the software and runs the configurePhase.

@@

Generally speaking, there are major gaps regarding the ability to successfully package the software between each different software solution that can be traced back to the build system.
Custom build systems require an disproportional amount of additional work to make them runnable on any system.
OpenFOAM and Allwmake are prime examples of these issues because now package managers have to understand a whole new build system for exactly one package in order to successfully package it and make it easily available to users.
It is also worth mentioning that writing your own custom build system, requires additional time to maintain this build system, that could have been spent working on the software itself.
Packages that use common, industry proven package managers, such as CMake and autotools make it easier for developers of the software to build their software as well as package managers to package it.
These build tools, generally also provide interfaces for dependency management that can be used to tell users and package maintainers what dependencies are needed, and it is usually better than only documenting required dependencies, because documentation needs to be kept in sync with the underlying code.
They also generally already provide proven interfaces for configuring packages with additional features, and these also might be self documenting.

The authors of the xSDK\footnote{\url{https://xsdk.info/}} came to a very similar conclusion.
Their goal is to ``identify, adapt, and adopt best practices in software engineering''~\cite{xsdk-website} and ``to provide the foundation of this extensible scientific software ecosystem''~\cite{xsdk-website}.
To achieve this they provide a list of package policies~\cite{xSDK2023}, that for example specifies as mandatory policy that packages ``must support portable installation through Spack''~\cite{xSDK2023} and that all packages ``should have a build system that is appropriate for the language''~\cite{xSDK2023}, that includes CMake and autoconf as examples.
If a package supports all these policies, they can be added to the growing list of packages in the xSDK, that currently among other things include PETSc, deal.ii and preCICE.
These packages are already part of nixpkgs, or it is easy to provide a package for them because they support portable installation, meaning installation into custom prefixes such as nix store paths, and common build systems such as CMake and autoconf.

@@

This chapter compares Nix with two widely used solution in HPC environments, Spack and EasyBuild.
There was a similar comparison done at FOSDEM 2018~\cite{comparision-fosdem} that also included conda and Guix, but this was by the release manager of EasyBuild and we want to provide an update with this comparison.
The comparison first takes a high level look on some statistics and how they compare and after that we will discuss, usability, reproducibly and performance in more detail.

@@

The first table~\ref{table:comparison-overview} shows some basic stats about the amount of packages the solutions provide out of the box and how they are implemented.
This is important because the higher the number the more likely it is that your package is already packaged and does not require additional work.
It is also important to look how packages are implemented because, if a package needs to be packaged by for example researches, domain specific languages have the downside that, researches now have to get familiar with this DSL, while general purpose languages such as python do not have this downside because they are already used in their field.
Although this is a downside, DSL usually are design with a purpose in mind and can get you faster to the desired outcome.

@@

Spack's way of installing has some problems, with using \texttt{git clone} on the latest default branch, in their case \texttt{develop}, this has the implication, that you are not using the latest stable release of the software but rather a random commit, that might have broken something recently.
It also has the problem that you cannot easily update the package repository independent from the package manager, this requires additional work from the users if they require this.

It is also worth mentioning that all 3 solutions do not require root for installation but Nix has the option for a system wide installation of the Nix daemon, that can then be used to install packages locally into the users PATH, without requiring root permissions.

@@

The basic usage of all three tools is pretty similar and they support searching and installing package in a simple manner, all this is show in table~\ref{table:comparison-usage}.
The difference in installing software comes down to the unique way that Nix installs software, it downloads/builds the paths into its local store, that can either be located in the user's home directory or in the path \texttt{/nix/store} depending on the installation.
\texttt{nix-shell} later only puts these downloaded paths into the \texttt{\$PATH} environment variable.
Spack and EasyBuild on the other hand use the widely used \texttt{module} system that usually is already found on HPC clusters, it allows loading software into the current environment.

Installing old software also differs for all three solutions, Nix on the one hand keeps some old, widely used software packages around and maintains those, examples are \texttt{boost} that is available in different versions or e.g. \texttt{hdf5} that is available in the latest version and version \texttt{1.10}.
The other way of installing older software on nix is using something that is called \texttt{overrideAttrs}, it can be used to change anything of a package, such as the used sources, meaning it can be used to replace it with an older/newer version or it can be used to add an patch, add a new dependency or install additional files into the \texttt{\$out} directory.
An example would be the following \texttt{nix-shell -p 'precice.overrideAttrs(\_: \{ version = "2.4.0"; src = \ldots \} )'}, it replaces the preCICE version with 2.4.0 and adds the old source, that will then trigger a new build.
Spack and EasyBuild on the other hand, keep the old software versions in the specifications or keep the whole specification, so you can reinstall an old version by specifying the available version, for example, \texttt{spack install gcc@6.4.0}
If the required version is not available, you have to edit the current package definition and add the required version.

To achieve reproducibility, it is necessary that we can install software from a previous revision of the packages.
Nix allows us to simply specify the channel revision we want to use to build our software, it then evaluates the requested packages and builds it.
This can easily be done without affecting the current system by specifying it in the shell command \texttt{nix-shell -p pan -I nixpkgs=pathToRevision/urlToTarball}, it now only uses this channel for building the requested package.
Spack, can achieve something similar by checking out your Spack version at a previous revision and then installing the package requested package, but this approach has a downside that it affects the whole Spack installation, and now also work with an older Spack tool.
On the other hand, EasyBuild, could be installed using pip at an earlier version, but it also offers something different called toolchains, we will look at shortly.

Next, we look at how you can build a package with a specific compiler version using specific compiler flags.
The flexibility of Nix overrides allows different levels of rebuilds of packages with a specific compiler, you can either just rebuild the requested package with specific compiler, this for example rebuilds curl with clang8 \texttt{nix-shell -p '(curl.override \{ stdenv = clang8Stdenv;\})'} or you can rebuild the package and all required dependencies with a specific compiler by overriding the \texttt{stdenv} on nixpkgs level, \texttt{nix-build -E 'with (import <nixpkgs> \{ config.replaceStdenv = \{ pkgs \}: pkgs.clang8Stdenv; \}); curl'}.
This is possible because every c/c++/\ldots package, is build using the \texttt{stdenv} that defines the compiler and default compile flags.
While Nix provides a default stdenv, that is used to compile all packages, it is possible to override and replace the package with your another stdenv provided by nixpkgs, such as the clang8Stdenv, you can write your own stdenv, or you can extend the current stdenv with flags.
By default, there are some useful helpers provided, such as \texttt{makeStaticLibraries} that allows compiling any library as a static library, or \texttt{enableDebugging} that sets the compiler flags \texttt{-ggdb -Og} and can then be enabled on the package plus all dependencies.
Spack provides an easy way to specify the compiler using \texttt{spack install precice \%clang} and only requires a registered compiler, it can be a system compiler, or the compiler can be installed using Spack.
On this note, Spack tooling and interface to override compilers, flags and more is way more user friendly and readable than using and invoking the Nix DSL within a one line command call.
EasyBuild has a whole different system, where packages have defined toolchains.\footnote{\url{https://docs.easybuild.io/common-toolchains/}}
A toolchain is a set of packages defined at a specific version and users can create their own toolchains.
For example, EasyBuild provides a foss toolchain that contains packages such as, binutils, gcc, OpenMPI, openblas and more and packages are provided against a specific toolchain version such as foss-2022b.
Packages are then provided with toolchains in mind and are never deleted, so installing a package with a different compiler can be achieved by selecting a different eb script with a different toolchain version.
On the other hand this results in a less elegant user flow, because if a user wants to install a package with a toolchain that is not provided they have to fork a package definition and update the toolchain.\\

@@

Spack does something similar to Nix where it adds hashes to each installation path, but also stores additional files inside the package installation including the complete spec of the package than can later be used to reproducibly build the same package.~\cite{Gamblin_2015}.
It also does something that Nix, currently lacks support for, namely storing the architecture and cpu features that are used in this build.
That way it can provide binary packages that are optimized for specific CPUs.
None the less it has one problem that Nix does not have: it allows packages to link back into system dependencies, for example, while building hpl-2.3 using a gcc 12.2.0 compiled with Spack we see the usage of the following system dependencies \texttt{libm.so}, \texttt{libc.so}, \texttt{libdl.so}, \texttt{libnl-3.so}, \texttt{libnl-route-3.so}, \texttt{librt.so}, \texttt{libutil.so}, \texttt{libpthread.so} as well as the dynamic loader.
Most of these dependencies are stable, and usually do not change on a system update, but it also means that it simply cannot guarantee the best possible reproducibility and it gets worse if the system compiler was used rather than a compiler build from scratch using Spack.
It is also worth mentioning that the Spack developers are aware of this problem and are currently working on building these system libraries themselves~\footnote{\url{https://github.com/spack/spack/issues/39560}}~\footnote{\url{https://github.com/spack/spack/pull/42082}}.

EasyBuild does achieve reproducibility with its toolchains and the approach to keep old package definitions, meaning that they also keep old dependency definitions and each dependency is pinned at a specific version~\cite{Geimer_2014}.
Regarding, CPU architecture and features, EasyBuild does not do generic builds by default but rather prioritizes the idea of optimizing for a specific system, meaning defaulting to \texttt{-march=native} for all packages and it has to manually be disabled~\footnote{\url{https://docs.easybuild.io/controlling-compiler-optimization-flags/\#controlling\_compiler\_optimization\_flags\_optarch\_default}}.
This leads to inherently impure and irreproducible builds as building the same easyconfig on two different host/architectures results in different binaries.
On the other hand, the already mentioned option for \texttt{osdependencies} goes against the reproducibility idea, because any os dependency is not managed by the package manager and might change at any point.
This leads to the problem of binaries linking back to the operating system \texttt{/usr/lib}, that Spack also has.
For instance, a curl build, uses the libraries libssl, libcrypto, libz, libm and libc, all of them being provided by the operating system, and for EasyBuild it includes libraries that change more frequently such as libssl, libcrypto and libz.\\
